BOLD Parameter Estimation Using Particle Filters

### 2010-08-15 doc/thesis/results.tex
mostly finished with simulation results

\subsection{Ideal FMRI Run}
\label{sec:VeryLongSim}

@@

\begin{figure}
\subfigure[Converging histogram for $q$, $s$, $f$, and $BOLD$ of the first run, low noise simulation.]
{\includegraphics[clip=true,trim=7cm 3cm 6cm 3cm, width=16cm]{images/converge_lownoise3}}
\caption{Convergence of the first run from \autoref{tab:LowNoiseResults}. The bars represent
a histogram, where darker bars indicate more particles in that bin.}
\label{fig:LowNoiseHist}
\end{figure}

@@

%HIGH NOISE SECTION, with a signal
\subsection{High Noise}
\label{sec:SimHighNoise}
%\begin{figure}
%\includegraphics[clip=true,trim=6cm 2cm 5cm 3.5cm,width=15cm]{images/realization_highnoise}
%\caption{Test Signals with high noise ($\sigma_x = 0.01, \sigma_y=0.005$) compared to the clean signal.}
%\label{fig:HighNoiseRealization}
%\end{figure}

@@

\begin{figure}[H]
\subfigure[$q$, $s$, $f$, $BOLD$, Run 1 ]
{\includegraphics[clip=true,trim=6cm 2cm 6cm 3cm, width=14cm]{images/highnoise_run5_3}}
\caption{Converging histogram for parameters during run 1, as in \autoref{fig:NoiseComparisonJustTwo}.}
\label{fig:ConvergenceRuns1}
\end{figure}

@@

\begin{figure}[H]
\subfigure[$q$, $s$, $f$, $BOLD$, Run 2 ]
{\includegraphics[clip=true,trim=6cm 2cm 6cm 3cm, width=14cm]{images/highnoise_run6_3}}
\caption{Converging histogram for parameters during run 2, as in \autoref{fig:NoiseComparisonJustTwo}.}
\label{fig:ConvergenceRuns2}
\end{figure}

The particles converged much faster when more noise was present (\autoref{fig:LowNoiseHist} vs. 
\autoref{fig:ConvergenceRuns1}, \autoref{fig:ConvergenceRuns2}). 
This also caused significantly more resampling which 
is the explanation for the percieved jumps in the histograms. 
The parameters arrived at for all 11 runs are shown in \autoref{tab:HighNoiseResults}.
Clearly the additional noise resulted in much more sporadic results. This is often the 
result when the particle filter converges too fast, in this case the result of the 
weighting function's variance being smaller than the measurement noise ($0.005$ vs. $0.01$). 
The error clearly suffers due to this effect (\autoref{tab:HighNoiseResults}).

@@

%NO SIGNAL, VERY HIGH NOISE
\subsection{Pure Noise, High Magnitude}
\label{sec:PureNoiseHighMag}
To determine how the particle filter responds to active, yet unrelated
portions of the brain, this section repeats the previous with much higher
noise peaks. To simulate this case another
pure noise signal was generated using $\sigma_x$ of $.1$ and $\sigma_y$ of $0.05$.

As before, the convergence all follows a similar path, leaving almost no
variance in the estimated time series (\autoref{fig:fits_noiseonly_high}). 
Interestingly the algorithm suffered from almost constant particle deprivation, 
meaning that the heuristic
for rescuing the particle filter from particle deprivation, discussed in
\autoref{sec:Resampling}, in fact was working against the proper course of action. 
The proper course of action in this case would be for the particle filter to fail,
since no particle can really properly estimate a random sequence. When this mechanism
was removed, all 11 runs stopped due to particle deprivation (all weights hit zero). 
The problem with allowing particle deprivation to occur is that it can \emph{rarely}
occur if the prior happens to miss the solution.

@@

Because of the preprocessing steps, the pure noise signal can look markedly like a real
signal (\autoref{fig:justbignoise_fit_0}). The preprocessing also causes the particle filter
to converge to a non-zero response in spite of the fact that the input does not correlate
with the stimuli in any way (\autoref{fits_noiseonly_high}).

\subsection{Single Voxel Review}
\label{sec:SingleVoxelReview}

The first step in determining the validity of a model is to provide some order of quality
to rate the results by. Unfortunately, because of the variability in the signal levels the
raw residual cannot perform this task. As demonstrated by \autoref{sec:PureNoiseLowMag},
a low residual does not necessarily indicate a good fit. 
Therefore, instead I  normalized the signal based on a value proportional to
the magnitude of the signal. Considering the tendency of FMRI noise to have large unexplainable
peaks and troughs, rather than using MAX and MIN values to estimate the scale of the 
signal, I used a robust estimator of 
scale; the Median-Absolute-Deviation (MAD). This is an estimator of the standard
deviation, and thus a good estimator of the scale of the input signal. 
The normalized residual values are 
shown in \autoref{tab:SignleVoxelActivationComparison}. A second potential method of 
gauging performance is mutual information. Mutual information is a method of measuring
the interdependence of two random variables. If two signals are truly independent,
then the mutual information will be zero. Although ideally suited to Bernoulli 
random variables where there are a finite number of options, by using histograms
it is possible to derive a joint distribution of two signals. The algorithm for 
mutual information is based on that joint distribution:
\begin{equation}
\sum_{x,y} p(x,y) \log_2\left(\frac{p(x,y)}{p(x)p(y)}\right)
\end{equation}
Unfortunately the number of bins causing bias in the output, thus to correct for this,
I subtracted the estimated bias:
\begin{equation}
\text{bias} = \frac{N_{bins}}{2Nlog(2)}
\end{equation}
where $N$ is the number of samples and $N_{bins}$ is the number of bins. For all
the mutual information estimates in this work 6 bins are used for the marginal
distribution of each signal. Additionally, throughout log base 2 will be used.
This leads to 36 total bins in the joint, so the bias is: 
\begin{equation}
\text{bias} = \frac{18}{N}
\end{equation}
After calculating the empirical mutual information, I subtracted the bias from
the estimated mutual information. Note that this can result in negative mutual
information, which should not technically be possible; so any negative mutual information
was taken as 0.

\begin{table}[t]
\centering
\begin{tabular}{|c | c | c | c | c | c | c | c | c |}
\hline 
& \multicolumn{4}{|c|}{Signal} & \multicolumn{4}{|c|}{No Signal}\\
\hline
& \multicolumn{2}{|c|}{Low Noise} & \multicolumn{2}{|c|}{High Noise} 
& \multicolumn{2}{|c|}{$\sigma_y = 0.001, \sigma_x = 0.0005$} 
& \multicolumn{2}{|c|}{$\sigma_y = 0.01, \sigma_x = 0.005$}\\
\hline 
& $M.I.$ & N. Res. &
  $M.I.$ & N. Res. &
  $M.I.$ & N. Res. &
  $M.I.$ & N. Res. \\
\hline       
\hline       
1 &   0.86687  &0.47801 &  0.09077  &1.03894 &  0.06326  &  1.29501 & 0.03024  &1.33641 \\
2 &   0.93975  &0.53177 &  0.13767  &0.95165 &  -0.01075  & 1.30175 & -0.02677 &1.33667 \\
3 &   0.82382  &0.5458  &  0.13505  &0.99539 &  0.02345  &  1.26287 & -0.0111  &1.15957 \\
4 &   0.94661  &0.49824 &  0.04341  &1.16129 &  -0.00906  & 1.43196 & 0.00147  &1.09988 \\
5 &   0.94281  &0.46805 &  0.13718  &1.03972 &  0.00663  &  1.25664 & -0.00204 &1.20107 \\
6 &   0.92539  &0.459   &  0.12337  &1.00214 &  -0.00816  & 1.2708 &  0.01775  &1.04589 \\
7 &   0.98892  &0.46096 &  0.15381  &1.08847 &  0.02664  &  1.15441 & 0.03163  &1.20543 \\
8 &   0.98796  &0.51838 &  0.11325  &1.05962 &  0.03285  &  1.27456 & 0.01951  &1.1225 \\
9 &   0.8804   &0.5253  &  0.09669  &1.0157  &  0.01628  &  1.32024 & 0.01039  &1.08637 \\
10 &  0.88721   &0.49211 & 0.18339  &1.18996 &  0.00407  &  1.34456 & 0.00508  &1.22135 \\
11 &  0.96644   &0.49092 & 0.10949  &0.95368 &  0.03323  &  1.32522 & -0.01284 &1.11737 \\
\hline                                                                         
mean &0.92329  &0.49714 & 0.12037   &1.04514 &  0.01622  &  1.29437 & 0.00576  &1.17568 \\
\hline                                                                           
min &  0.82382  &0.459   &  0.04341  &0.95165 & -0.01075  & 1.15441 & -0.02677 &1.04589 \\
\hline                                                                         
max &  0.98892  &0.5458  & 0.18339   &1.18996 & 0.06326  &  1.43196 & 0.03163  &1.33667 \\
\hline
\end{tabular}
\caption{Mutual Information and the normalized $\sqrt{MSE}$, 
    for signal/noise configurations.}
\label{tab:SignleVoxelActivationComparison} 
\end{table}

Comparing the results of \autoref{sec:SimHighNoise} and \autoref{sec:PureNoiseHighMag}, 
distinguishing between these cases with ether normalized residual or mutual information
is not clear cut. While the average mutual information is in fact more than 10 times
greater than the average mutual information in the two non-signal cases, the maximum
mutual information of the low noise/no signal case exceeds the minimum M.I. of the high noise/signal
case. The Low Noise/ signal determination is easier to make; given
the minimum mutual information is above $.8$ and the maximum normalized residual is below $.6$.
However, it is worth noting that the worst case scenario  for mutual information (maximum) in the
low noise/no signal case does not concur with the worse case (minimum) normalized residual. 
There is no reason why this has to be the case, it is beneficial. In
other words, if it were necessary to make a statement that
a particular voxel active or inactive; the accuracy would improve if both techniques
were used with loose restrictions, but a voxel would only be considered active if the
voxel passed both tests.

There were two primary purposes of these tests. First, given the nature of monte-carlo
techniques it is important to ensure consistency of results. Although the parameter
sets were somewhat inconsistent, the quality of the fit was actually quite consistent
across eleven runs. The second purpose was of course to determine how the particle 
filter responded to different signal to noise ratios.

\section{Multi-voxel Simulation}
\label{sec:Multi-voxel Simulation}
To test the usefulness of the particle filter for I used a modified version of the FSL tool 
POSSUM to generate an entire FMRI image from a parameter map. The parameter map was generated
by taking an existing activation map and assigning discrete parameter sets to each region.
The result was a four dimensional (length x width
x height x parameter) image with spatially varying parameters. Possum was then modified
to take a parameter map and generate activation levels depending on the parameters at that
point. The patch for POSSUM will be made available. As an unfortunate side effect of 
not using Possums' original activation scale, I manually added 750 to the total level of
simulated Possum images. This is because the BOLD \% difference levels were in the range
of 50 - 100\% from the base, about 5 times as large as they should have been. Ultimately
this has no effect on the quality of the simulation and should not effect the 
underlying parameters (other than perhaps $\epsilon$ and $V_0$). 
Also worth noting is that the noise level was set to an SNR of 20, due to changes in the 
program the true signal-to-noise ratio was much lower, as seen in the fit in \autoref{fig:simslicefit1}.

\begin{figure}
\centering
\includegraphics[width=15cm]{images/simslice_noise}
\caption{A single time-series fit in the simulated slice.}
\label{fig:simslicefit1}
\end{figure}

For each time-series in the simulated FMRI image, the final \emph{static} parameters were saved
into a parameter map. This parameter map may then be compared to the map used to generate the 
simulated data; additionally a new simulation using the calculated parameters may also be 
generated to test the difference in BOLD levels between the real parameters and the
estimated ones. Since it is the parameters are far from orthogonal 
(\cite{Deneux2006}), its possible that two sets of parameters are functionally equivalent,
despite large difference. This way, a quantitative difference between the two parameter sets 
was found.

@@

The histograms again demonstrate that a single point estimate of the parameters 
is elusive for the current model. However the data clearly show the power
of the particle filter at identifying regions of activation; which is usually
performed with statistical parametric maps. The thresholds applied to this
slice, both for mutual information and residuals is arbitrary, and needs further
research. Tighter thresholds removes the false positives present
in the images at the cost of false negatives. Interestingly the false 
positives present in the Mutual Information
map are different from those in the residual map. This furthers the argument from
\autoref{sec:SingleVoxelReview} for combining the two metrics to 
increase power. Although
at first glance it would appear that there are false negatives in the 
\autoref{fig:simslice_hm}; this is
not actually the case. POSSUM simulates different tissues, and white matter
does not typically  have a BOLD response. This is why there are holes in regions
2 and 3. These results certainly indicate that the particle filter is effective
at regressing against a noisy signal. Therefore the next section moves forward
with real FMRI data.

>>>>

\subsection{Ideal Analysis}
\label{sec:VeryLongSim}

@@

\begin{figure}
\subfigure[Converging histogram for $q$, $s$, $f$, and $BOLD$ of the first run, low noise simulation.]
{\label{fig:LowNoiseHistc} \includegraphics[clip=true,trim=7cm 3cm 6cm 3cm, width=16cm]{images/converge_lownoise3}}
\caption{Convergence of the first run from \autoref{tab:LowNoiseResults}. The bars represent
a histogram, where darker bars indicate more particles in that bin.}
\label{fig:LowNoiseHist}
\end{figure}

@@

%HIGH NOISE SECTION, with a signal
\subsection{Simulation with High Noise}
\label{sec:SimHighNoise}
%\begin{figure}
%\includegraphics[clip=true,trim=6cm 2cm 5cm 3.5cm,width=15cm]{images/realization_highnoise}
%\caption{Test Signals with high noise ($\sigma_x = 0.01, \sigma_y=0.005$) compared to the clean signal.}
%\label{fig:HighNoiseRealization}
%\end{figure}

@@

\begin{figure}[H]
\subfigure[$q$, $s$, $f$, $BOLD$, Run 1 ]
{\label{fig:ConvergenceRuns1c} \includegraphics[clip=true,trim=6cm 2cm 6cm 3cm, width=14cm]{images/highnoise_run5_3}}
\caption{Converging histogram for parameters during run 1, as in \autoref{fig:NoiseComparisonJustTwo}.}
\label{fig:ConvergenceRuns1}
\end{figure}

@@

\begin{figure}[H]
\subfigure[$q$, $s$, $f$, $BOLD$, Run 2 ]
{\label{fig:ConvergenceRuns2c} \includegraphics[clip=true,trim=6cm 2cm 6cm 3cm, width=14cm]{images/highnoise_run6_3}}
\caption{Converging histogram for parameters during run 2, as in \autoref{fig:NoiseComparisonJustTwo}.}
\label{fig:ConvergenceRuns2}
\end{figure}

The particles converged much faster when more noise was present (\autoref{fig:LowNoiseHist} vs. 
\autoref{fig:ConvergenceRuns1}, \autoref{fig:ConvergenceRuns2}). 
This also caused significantly more resampling which 
is the explanation for the perceived jumps in the histograms. 
Clearly the additional noise resulted in much more sporadic results (\autoref{tab:HighNoiseResults}). 
This is often the 
result when the particle filter converges too fast, in this case the result of the 
weighting function's variance being smaller than the measurement noise ($0.005$ vs. $0.01$). 
The error clearly suffers due to this effect (\autoref{tab:HighNoiseResults}).
Note that neither Run 1 or Run 2 estimated the underlying state well (\autoref{fig:ConvergenceRuns1c}
and \autoref{fig:ConvergenceRuns2c}), whereas
in the previous test the particle filter was extremely successful in this area (\autoref{fig:LowNoiseHistc}).

@@

%NO SIGNAL, VERY HIGH NOISE
\subsection{Pure Noise, High Magnitude}
\label{sec:PureNoiseHighMag}
To determine how the particle filter responds to active, yet unrelated
portions of the brain, this section repeats the test of \autoref{sec:PureNoiseLowMag} 
with much higher noise peaks. To simulate this case another
pure noise signal was generated using $\sigma_x$ of $0.1$ and $\sigma_y$ of $0.05$.

As before, the convergence all follows a similar path, leaving almost no
variance in the estimated time series (\autoref{fig:fits_noiseonly_high}). 
Interestingly the algorithm suffered from almost constant particle deprivation, 
meaning that the heuristic
for rescuing the particle filter from particle deprivation, discussed in
\autoref{sec:Resampling}, in fact was working against the proper course of action. 
The proper course of action in this case would be for the particle filter to fail,
since no particle can really properly estimate a random sequence. When this mechanism
was removed, all 11 runs stopped due to particle deprivation (all weights hit zero). 
The problem with allowing particle deprivation to occur is that it can \emph{rarely}
occur in good data if the prior distribution does not cover a solution.

Because of the preprocessing steps, the pure noise signal can look markedly like a real
signal (\autoref{fig:justbignoise_fit_0}). The preprocessing causes the particle filter
to converge to a non-zero response in spite of the fact that the input does not correlate
with the stimuli in any way (\autoref{fig:fits_noiseonly_high}).

@@

\subsection{Single Voxel Review}
\label{sec:SingleVoxelReview}
The first step in determining the validity of a model is to provide some order of quality
to rate the results by. Unfortunately, because of the variability in the signal levels, the
raw residual cannot perform this task. As demonstrated by \autoref{tab:NoiseOnlyResults},
a low residual does not necessarily indicate a good fit. 
Therefore, instead I  normalized the signal based on a value proportional to
the magnitude of the signal. Considering the tendency of FMRI noise to have large unexplainable
peaks and troughs, rather than using MAX and MIN values to estimate the scale of the 
signal, I used a robust estimator of 
scale; the Median-Absolute-Deviation (MAD) which is defined in equation \autoref{eq:mad}.
This is an estimator of the standard
deviation, and thus a good estimator of the scale of the input signal. 
The normalized residual values are 
shown in \autoref{tab:SingleVoxelActivationComparison}. A second potential method of 
gauging performance is mutual information. Mutual information is a method of measuring
the interdependence of two random variables. If two signals are truly independent,
then the mutual information will be zero. Although ideally suited to discrete
distributions, by using histograms
it is possible to derive a joint distribution of two signals. The algorithm for 
mutual information is based on that joint distribution:
\begin{equation}
\sum_{x,y} p(x,y) \log_2\left(\frac{p(x,y)}{p(x)p(y)}\right)
\end{equation}
Unfortunately the number of bins causes bias in the output, thus to correct for this,
I subtracted the estimated bias:
\begin{equation}
\text{bias} = \frac{N_{bins}}{2Nlog(2)}
\end{equation}
where $N$ is the number of samples and $N_{bins}$ is the number of bins. For all
the mutual information estimates in this work 6 bins are used for the marginal
distribution of each signal. Additionally, throughout log base 2 will be used.
This leads to 36 total bins in the joint, so the bias is: 
\begin{equation}
\text{bias} = \frac{18}{N}
\end{equation}
 Note that subtracting the bias can result in negative mutual
information, which should not technically be possible; so any negative mutual information
was taken as 0.

\begin{table}[t]
\centering
\begin{tabular}{|c | c | c | c | c | c | c | c | c |}
\hline 
& \multicolumn{4}{|c|}{Signal} & \multicolumn{4}{|c|}{No Signal}\\
\hline
& \multicolumn{2}{|c|}{Low Noise} & \multicolumn{2}{|c|}{High Noise} 
& \multicolumn{2}{|c|}{$\sigma_y = 0.001, \sigma_x = 0.0005$} 
& \multicolumn{2}{|c|}{$\sigma_y = 0.01, \sigma_x = 0.005$}\\
\hline 
& $M.I.$ & N. Res. &
  $M.I.$ & N. Res. &
  $M.I.$ & N. Res. &
  $M.I.$ & N. Res. \\
\hline       
\hline       
1 &   0.86687  &0.47801 &  0.09077  &1.03894 &  0.06326  &  1.29501 & 0.03024  &1.33641 \\
2 &   0.93975  &0.53177 &  0.13767  &0.95165 &  -0.01075  & 1.30175 & -0.02677 &1.33667 \\
3 &   0.82382  &0.5458  &  0.13505  &0.99539 &  0.02345  &  1.26287 & -0.0111  &1.15957 \\
4 &   0.94661  &0.49824 &  0.04341  &1.16129 &  -0.00906  & 1.43196 & 0.00147  &1.09988 \\
5 &   0.94281  &0.46805 &  0.13718  &1.03972 &  0.00663  &  1.25664 & -0.00204 &1.20107 \\
6 &   0.92539  &0.459   &  0.12337  &1.00214 &  -0.00816  & 1.2708 &  0.01775  &1.04589 \\
7 &   0.98892  &0.46096 &  0.15381  &1.08847 &  0.02664  &  1.15441 & 0.03163  &1.20543 \\
8 &   0.98796  &0.51838 &  0.11325  &1.05962 &  0.03285  &  1.27456 & 0.01951  &1.1225 \\
9 &   0.8804   &0.5253  &  0.09669  &1.0157  &  0.01628  &  1.32024 & 0.01039  &1.08637 \\
10 &  0.88721   &0.49211 & 0.18339  &1.18996 &  0.00407  &  1.34456 & 0.00508  &1.22135 \\
11 &  0.96644   &0.49092 & 0.10949  &0.95368 &  0.03323  &  1.32522 & -0.01284 &1.11737 \\
\hline                                                                         
mean &0.92329  &0.49714 & 0.12037   &1.04514 &  0.01622  &  1.29437 & 0.00576  &1.17568 \\
\hline                                                                           
min &  0.82382  &0.459   &  0.04341  &0.95165 & -0.01075  & 1.15441 & -0.02677 &1.04589 \\
\hline                                                                         
max &  0.98892  &0.5458  & 0.18339   &1.18996 & 0.06326  &  1.43196 & 0.03163  &1.33667 \\
\hline
\end{tabular}
\caption{Mutual Information and the normalized $\sqrt{MSE}$, 
    for signal/noise configurations.}
\label{tab:SingleVoxelActivationComparison} 
\end{table}

Comparing the results of \autoref{sec:SimHighNoise} and \autoref{sec:PureNoiseHighMag} 
in \autoref{tab:SingleVoxelActivationComparison}, 
distinguishing between these cases with ether normalized residual or mutual information
is not clear cut. While the average mutual information is in fact more than 10 times
the average mutual information in the two non-signal cases, the maximum
mutual information of the low noise/no signal case exceeds the minimum M.I. of the high noise/signal
case. The Low Noise/ signal determination is easier to make; given
the minimum mutual information is above $.8$ and the maximum normalized residual is below $.6$.
However, it is worth noting that the worst case scenario  for mutual information (maximum) in the
low noise/no signal case does not concur with the worse case (minimum) normalized residual. 
There is no reason why this has to be the case, but it is beneficial. In
other words, if it were necessary to make a statement that
a particular voxel were active or inactive; the accuracy would improve if both techniques
were used with loose restrictions.

There were two primary purposes of these tests. First, given the nature of monte-carlo
techniques it is important to ensure consistency of results. Although the parameter
sets were inconsistent, the quality of the fit was actually quite consistent
over eleven runs. The second purpose was of course to determine how the particle 
filter responded to different signal to noise ratios. From \autoref{sec:SimLowNoise},
the particle filter performed extremely well in filtering out nonsensical measurements.

\section{Multi-voxel Simulation}
\label{sec:Multi-voxel Simulation}
To test the usefulness of the particle filter on a larger scale, I used a modified version 
of the FSL tool 
POSSUM to generate an entire FMRI image from a parameter map. The parameter map was generated
by taking an existing activation map and assigning discrete parameter sets to each region.
The result was a four dimensional (length x width
x height x parameter) image with spatially varying parameters. Possum was then modified
to take a parameter map and generate activation levels depending on the parameters at that
point. The patch for POSSUM will be made available. As side effect of 
not using POSSUM's original activation scale, the BOLD \% difference levels 
were in the range of 50 - 100\% from the base, about 5 times as large as they 
should have been. I therefore manually added 750 to the total level of
simulated POSSUM images, putting results into the 1 - 5\% range. This 
had no effect on the quality of the simulation and should not effect the 
underlying parameters (other than $V_0$). 
Also worth noting is that the noise level was set to an SNR of 20, due to changes 
in the program the true signal-to-noise ratio was much lower, as seen in the 
fit in \autoref{fig:simslicefit1}.

%todo where is this from?
\begin{figure}
\centering
\includegraphics[clip=true,trim=5cm 2cm 4.5cm 2cm,width=15cm]{images/simslice_noise}
\caption{A single time-series fit in the simulated slice. Preprocessed BOLD is
the noisy response generated by POSSUM.}
\label{fig:simslicefit1}
\end{figure}

For each time-series in the simulated FMRI image, the final parameters were saved
into a parameter map. This parameter map could then be compared to the map used to generate the 
simulated data; additionally a new simulation using the calculated parameters could also be 
generated to test the difference in BOLD levels between the real parameters and the
estimated ones. Since the parameters were far from orthogonal 
\cite{Deneux2006}, this provided a quantitative difference between the two parameter sets.

@@

%todo quantify the number of false positives, add extra Heatmap removing false positives
The histograms again demonstrate that a single point estimate of the parameters 
is elusive for this set of parameters. However the data clearly show the power
of the particle filter at identifying regions of activation; which is usually
performed with statistical parametric maps. The thresholds applied to this
slice, both for mutual information and residuals is arbitrary, and needs further
research. Tighter thresholds removes the false positives present
in the images at the cost of false negatives. As noted in \autoref{sec:SingleVoxelReview}
the false positives present in the Mutual Information
map are different from those in the residual map. This furthers the argument from
for combining the two metrics to increase power. Although
at first glance it would appear that there are false negatives in the 
\autoref{fig:simslice_hm}; this is
not actually the case. POSSUM simulates different tissues, and white matter
does not typically  have a BOLD response. This is why there are holes in regions
2 and 3. These results indicate that the particle filter is effective
at regressing against a noisy signal.

### 2010-08-15 doc/thesis/resultsB.tex
mostly finished with simulation results

%\chapter{Real Data}
%Finally, we also performed inference on a real FMRI scan. The scanner we used...
%... more specifics...

% TODO include single?
%Before performing tests on a full image, I the particle filter
%on regions deemed active and inactive by statistical parametric mapping
%(SPM). This served the purpose of adjusting the priors as well as the 
%preprocessing based on real world signals. This was actually done before
%the simulations, and then results were carried back the simulations 
%to check consistency.
%After work adjusting parameters, most importantly the weighting function and the 
%priors, particle filter was applied to every voxel in an FMRI image.
%The results of this large scale analysis was a parameter map which was
%then used to calculate normalized square-root MSE image. 
%
%\section{Single-Voxel Analysis}
%The choice of a prior, as discussed previously, is extremely important. While a
%prior may have the potential to give good results, being a monte-carlo algorithm
%there is the possibility for inconsistencies. Thus, increasing the variance
%of the time-constants may allow additional flexibility, it will also cause
%additional model variance. Before running on a full volume I adjusted the 
%prior to ensure that the same input would give the same output 100 times in a 
%row. While this may seem like a given, with a random drawing of the prior,
%this can be difficult. Case in point, the exact same algorithm run twice
%with the time constants all having standard deviations of $.35$ resulted in two
%very different fits, shown in \autoref{fig:badfit_param1}.
%
%\begin{figure}
%\subfigure{\includegraphics[clip=true,trim=6cm 2cm 6cm 3.5cm,width=17cm]{images/badfit_param1}}
%\subfigure{\includegraphics[clip=true,trim=6cm 2cm 6cm 3.5cm,width=17cm]{images/goodfit_param1}}
%\caption{The same priors gave rise to both fits.}
%\label{fig:badfit_param1}
%\end{figure}
%
%For this reason, I actually lowered the standard deviations of the time
%constants to prevent over-smoothing. This resulted in more consistent,
%though potentially slightly worse fits, two examples of which are 
%shown in \autoref{fig:param2_var}. 
%
%\begin{figure}
%\subfigure{\includegraphics[clip=true,trim=6cm 2cm 6cm 3.5cm,width=17cm]{images/param2a}}
%\subfigure{\includegraphics[clip=true,trim=6cm 2cm 6cm 3.5cm,width=17cm]{images/param2b}}
%\caption{A poor fit, using the same parameters as }
%\label{fig:param2_var}
%\end{figure}
%
%todo: stats of the 100 fits?
%
%\section{Single Time-Series Simulation}
%
\chapter{Real Data}
\label{sec:RealData}
Modeling the BOLD response is of course not of much use if does not work for
real FMRI.  Although this algorithm will hopefully lead to more novel methods 
of analysis, the standard use for modeling the BOLD signal is to locate activation.
Activation is defined as areas where the input is the primary drive
for the BOLD response, as opposed to intermediate factors controlling it. 
Once areas where the BOLD 
model may be accurately estimated are found, integrating the model will allow for accurate
estimation of the state between measurements, which could then be used for more 
advanced analysis, for instance of areas that are being driven by other brain regions.
It all begins with localizing the first activation regions in the
chain. Therefore this section compares the output of the particle filter 
with conventional SPM.

The data described in this section are fundamentally different in several 
ways. SPM preprocesses the image by spatially smoothing the FMRI image (in this section 
SPM8 was used with an $8mm x 8mm x 8mm$ Gaussian kernel), whereas
this is not done in the particle filter algorithm. Additionally, a spline
was used to de-trend, rather than SPM8's high pass filter (with a cut
off based on a globally estimated autocorrelation). Thus the preprocessing pipelines 
are different; but the output of SPM8 is also different. Whereas SPM outputs
a t-statistic for each voxel, the output 
of the particle filter is a posterior probability distribution of the parameters
at every voxel. To validate the quality of the particle filter results 
it is necessary to compare the both the location and the fit calculated by the particle filter
with SPM's location and fit.

\section{FMRI Configuration}
For the FMRI data discussed in \autoref{sec:RealData}, tests were 
performed on a right handed volunteer using a Gradient EPI sequence
on a 1.5T GE scanner with a 2.1s TR. Slices were 5mm thick and echo
time was (TODO). The subject was then presented with either a single or
double flash, at randomly dispersed time points over the course of
5 minutes, 10 seconds. The FMRI began 18.9 seconds
before the beginning of the experiment, to allow for transients 
in to image to settle out. The timing of the flashes are shown in 
\autoref{tab:flashes}, with time 0 corresponding to the beginning of
the 10th TR (so 9 images were dropped).

\begin{table}[t]
\centering
\begin{tabular}{|c || c |}
\hline 
Single Flash Times & Double Flash Times \\
\hline 
.7 & 2\\
2.7 & 21\\
5.7 & 28\\
12.7 & 52\\
\hline 
\end{tabular}
\caption{Timing of single and double flashes.}
\label{tab:flashes} 
\end{table}

After dropping the first 9 volume, each of the remaining volumes was
co-registered with the first volume in time. At this point the SPM method
diverges from the experimental method. For the particle filter, detrending 
was then applied as discussed in \autoref{sec:Detrend} and the resulting
data was processed with the particle filter. To generate the SPM output
in this section, the co-registered data was passed through a high pass
filter with a cut off of (TODO). The data was then smoothed spatially 
with a Gaussian kernel (FWHM 8x8x8 voxels). A Statistical Parametric Map
was then generated using SPM8. All settings used in SPM8 were the defaults
for analyzing FMRI data.

>>>>

\chapter{Real Data}
\label{sec:RealData}
The ultimate goal of modeling the BOLD response is to estimate parameters
in real FMRI data. The most basic use of modeling the BOLD signal is to
locate activation. A voxel is considered active when the stimulus 
is the primary drive for the BOLD response. This is in contrast to 
most voxels which have are controlled by intermediate or complete unrelated factors. 
Inactive regions cannot be modeled because their input is unknown, therefore
parameter estimates in such regions are impossible. 
Because SPM is the de facto standard for localizing activation,
this section compares its output with that of the particle filter.

%todo check SPM smoothing parameter
Note that SPM must pre-process the image with a spatial smoothing 
filter. For this work SPM8 was smoothed with a
$8\text{mm} \times  8\text{mm} \times 8\text{mm}$ Full-Width Half Maximum
(FWHM) Gaussian kernel.  
Additionally, SPM8 applied a high pass filter (with a cut
off based on a globally estimated autocorrelation). Thus the preprocessing pipeline
of SPM is very different from that of the particle filter. SPM8 also outputs
a T-statistic for each voxel, whereas the 
particle filter's primary output is a posterior probability distribution of the parameters
at every voxel. To validate the quality of the particle filter, the results 
were compared with SPM, both in terms of the location and the fit.

%todo check scanner information/stimulus
\section{FMRI Configuration}
For the FMRI data discussed in \autoref{sec:RealData}, tests were 
performed on a right handed volunteer using a Gradient EPI sequence
on a 1.5T GE scanner with a 2.1s TR. Slices were 5mm thick and echo
time was (TODO). The subject was then presented with either a single or
double flash, at randomly dispersed time points over the course of
5 minutes, 10 seconds. The FMRI began 18.9 seconds
before the beginning of the experiment, to allow for transients 
in to image to settle out. The timing of the flashes are shown in 
\autoref{tab:flashes}, with time 0 corresponding to the beginning of
the 10th TR (so 9 images were dropped).

%todo the stimulus
\begin{table}[t]
\centering
\begin{tabular}{|c || c |}
\hline 
Single Flash Times & Double Flash Times \\
\hline 
.7 & 2\\
2.7 & 21\\
5.7 & 28\\
12.7 & 52\\
\hline 
\end{tabular}
\caption{Timing of single and double flashes.}
\label{tab:flashes} 
\end{table}

%spm options
After dropping the first 9 volumes, each of the remaining volumes was
co-registered with the first. At this point the SPM method
diverges from the experimental method. For the particle filter, detrending 
was then applied as discussed in \autoref{sec:Detrend} and the resulting
data was processed with the particle filter. To generate the SPM output
in this section, the co-registered data was spatially smoothed,
and then filtered with an adaptive cut off (which is built into the
SPM analysis). For the SPM analysis, the Canonical HRF was used
and model derivatives were included. All settings used in SPM8 were the defaults
for analyzing FMRI data.
